---
title: "Hands-on Exercise 2"
description: ""
date: "April 24, 2025"
date-modified: "April 25, 2025"
format: html
author: "Wang Shen Si"
editor: visual
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# **Beyond ggplot2 Fundamentals**

## **Overview**

In this chapter, you will be introduced to several ggplot2 extensions for creating more elegant and effective statistical graphics. By the end of this exercise, you will be able to:

-   control the placement of annotation on a graph by using functions provided in ggrepel package,

-   create professional publication quality figure by using functions provided in ggthemes and hrbrthemes packages,

-   plot composite figure by combining ggplot2 graphs by using patchwork package.

## Getting Started

### **Installing and loading the required libraries**

In this exercise, beside tidyverse, four R packages will be used. Code chunk below will be used to check if these packages have been installed and also will load them onto your working R environment.

```{r}
pacman::p_load(ggrepel, patchwork, ggthemes, hrbrthemes, tidyverse) 
```

### **Importing data**

The code chunk below imports *exam_data.csv* into R environment by using [*read_csv()*](https://readr.tidyverse.org/reference/read_delim.html) function of [**readr**](https://readr.tidyverse.org/) package. **readr** is one of the tidyverse package.

```{r}
exam_data <- read_csv("data/Exam_data02.csv", show_col_types = FALSE)
```

## **Beyond ggplot2 Annotation: ggrepel**

One of the challenge in plotting statistical graph is annotation, especially with large number of data points.

The Plot:

```{r}
#| echo: false
#| warning: false
ggplot(data = exam_data, 
       aes(x = MATHS, y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm, size = 0.5) +  
  geom_label(aes(label = ID), 
             hjust = .5, 
             vjust = -.5) +
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

The Code:

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label(aes(label = ID), 
             hjust = .5, 
             vjust = -.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

### **Working with ggrepel**

The Plot:

```{r}
#| echo: false
#| warning: false
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label_repel(aes(label = ID), 
                   fontface = "bold") +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

The Code:

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label_repel(aes(label = ID), 
                   fontface = "bold") +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

## **Beyond ggplot2 Themes**

ggplot2 comes with eight [built-in themes](https://ggplot2.tidyverse.org/reference/ggtheme.html), they are: , , , , , , , and .`theme_gray()theme_bw()theme_classic()theme_dark()theme_light()theme_linedraw()theme_minimal()theme_void()`

The Plot:

```{r}
#| echo: false
#| warning: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_gray() +
  ggtitle("Distribution of Maths scores") 
```

```{r}
#| eval: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_gray() +
  ggtitle("Distribution of Maths scores") 
```

Refer to this [link](https://ggplot2.tidyverse.org/reference/index.html#themes) to learn more about ggplot2 `Themes`

### **Working with ggtheme package**

[**ggthemes**](https://cran.r-project.org/web/packages/ggthemes/index.html) provides [‘ggplot2’ themes](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/) that replicate the look of plots by Edward Tufte, Stephen Few, [Fivethirtyeight](https://fivethirtyeight.com/), [The Economist](https://www.economist.com/graphic-detail), ‘Stata’, ‘Excel’, and [The Wall Street Journal](https://www.pinterest.com/wsjgraphics/wsj-graphics/), among others.

In the example below, *The Economist* theme is used.

```{r}
#| echo: false
#| warning: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_economist()
```

The Code:

```{r}
#| eval: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_economist()
```

It also provides some extra geoms and scales for ‘ggplot2’. Consult [this vignette](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/) to learn more.

### **Working with hrbthems package**

[**hrbrthemes**](https://cran.r-project.org/web/packages/hrbrthemes/index.html) package provides a base theme that focuses on typographic elements, including where various labels are placed as well as the fonts that are used.

The Plot:

```{r}
#| echo: false
#| warning: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum()
```

The Code:

```{r}
#| eval: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum()
```

The second goal centers around productivity for a production workflow. In fact, this “production workflow” is the context for where the elements of hrbrthemes should be used. Consult [this vignette](https://hrbrmstr.github.io/hrbrthemes/articles/why_hrbrthemes.html) to learn more.

The Plot:

```{r}
#| echo: false
#| warning: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```

The Code:

```{r}
#| eval: false
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```

## **Beyond Single Graph**

It is not unusual that multiple graphs are required to tell a compelling visual story. There are several ggplot2 extensions provide functions to compose figure with multiple graphs. In this section, you will learn how to create composite plot by combining multiple graphs. First, let us create three statistical graphics by using the code chunk below.

The Plot:

```{r}
#| echo: false
#| warning: false
p1 <- ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") + 
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of Maths scores")
p1
```

The Code:

```{r}
#| eval: false
p1 <- ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") + 
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of Maths scores")
```

Next

The Plot:

```{r}
#| echo: false
#| warning: false
p2 <- ggplot(data=exam_data, 
             aes(x = ENGLISH)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of English scores")
p2
```

The Code:

```{r}
#| eval: false
p2 <- ggplot(data=exam_data, 
             aes(x = ENGLISH)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of English scores")
```

Lastly, we will draw a scatterplot for English score versus Maths score by as shown below

The Plot:

```{r}
#| echo: false
#| warning: false
p3 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
p3
```

The Code:

```{r}
#| eval: false
p3 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

### **Combining two ggplot2 graphs**

Figure in the tabset below shows a composite of two histograms created using patchwork. Note how simple the syntax used to create the plot!

The Plot:

```{r}
#| echo: false
#| warning: false
p1 + p2
```

The Code:

```{r}
#| eval: false
p1 + p2
```

### **Combining three ggplot2 graphs**

We can plot more complex composite by using appropriate operators. For example, the composite figure below is plotted by using:

-   “/” operator to stack two ggplot2 graphs,

-   “\|” operator to place the plots beside each other,

-   “()” operator the define the sequence of the plotting.

The Plot:

```{r}
#| echo: false
#| warning: false
(p1 / p2) | p3
```

The Code:

```{r}
#| eval: false
(p1 / p2) | p3
```

To learn more about, refer to [Plot Assembly](https://patchwork.data-imaginist.com/articles/guides/assembly.html).

### **Creating a composite figure with tag**

In order to identify subplots in text, **patchwork** also provides auto-tagging capabilities as shown in the figure below.

The Plot:

```{r}
#| echo: false
#| warning: false
((p1 / p2) | p3) + 
  plot_annotation(tag_levels = 'I')
```

The Code:

```{r}
#| eval: false
((p1 / p2) | p3) + 
  plot_annotation(tag_levels = 'I')
```

### **Creating figure with insert**

Beside providing functions to place plots next to each other based on the provided layout. With [`inset_element()`](https://patchwork.data-imaginist.com/reference/inset_element.html) of **patchwork**, we can place one or several plots or graphic elements freely on top or below another plot.

The Plot:

```{r}
#| echo: false
#| warning: false
p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1)
```

The Code:

```{r}
#| eval: false
p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1)
```

### **Creating a composite figure by using patchwork and ggtheme**

Figure below is created by combining patchwork and theme_economist() of ggthemes package discussed earlier.

The Plot:

```{r}
#| echo: false
#| warning: false
patchwork <- (p1 / p2) | p3
patchwork & theme_economist()
```

The Code:

```{r}
#| eval: false
patchwork <- (p1 / p2) | p3
patchwork & theme_economist()
```

## **Reference**

-   [Patchwork R package goes nerd viral](https://www.littlemissdata.com/blog/patchwork)

-   [ggrepel](https://ggrepel.slowkow.com/)

-   [ggthemes](https://ggplot2.tidyverse.org/reference/ggtheme.html)

-   [hrbrthemes](https://cinc.rud.is/web/packages/hrbrthemes/)

-   [ggplot tips: Arranging plots](https://albert-rapp.de/post/2021-10-28-extend-plot-variety/)

-   [ggplot2 Theme Elements Demonstration](https://henrywang.nl/ggplot2-theme-elements-demonstration/)

-   [ggplot2 Theme Elements Reference Sheet](https://isabella-b.com/blog/ggplot2-theme-elements-reference/)
