---
title: "Hands-on Exercise 1"
description: ""
date: "April 19, 2025"
date-modified: "April 22, 2025"
format: html
author: "Wang Shen Si"
editor: visual
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Getting Started

## Install and launching R packages

The code chunk below uses p_load() of pacman package to check if tidyverse packages are installed in the computer. If they are, then they will be launched into R.

```{r}
pacman::p_load(tidyverse)
```

## Importing the data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

# Plotting a simple bar chart

```{r}
ggplot(data = exam_data,
       aes(x = RACE)) +
  geom_bar()
```

R Graphics VS ggplot

First, let us compare how R Graphics, the core graphical functions of Base R and ggplot plot a simple histogram.

R Graphics:

```{r}
hist(exam_data$MATHS)
```

ggplot:

```{r}
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram(bins=10, 
                 boundary = 100,
                 color="black", 
                 fill="grey") +
  ggtitle("Distribution of Maths scores")
```

## Geometric Objects: geom_bar

The code chunk below plots a bar chart by using [`geom_bar()`](https://ggplot2.tidyverse.org/reference/geom_bar.html).

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```

## Geometric Objects: geom_dotplot

In the code chunk below, [`geom_dotplot()`](https://ggplot2.tidyverse.org/reference/geom_dotplot.html) of ggplot2 is used to plot a dot plot.

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(dotsize = 0.5)
```

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,         
               dotsize = 0.5) +      
  scale_y_continuous(NULL,           
                     breaks = NULL)  
```

## Geometric Objects: `geom_histogram()`

In the code chunk below, [*geom_histogram()*](https://ggplot2.tidyverse.org/reference/geom_histogram.html) is used to create a simple histogram by using values in *MATHS* field of *exam_data*.

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram()       
```

## Modifying a geometric object by changing `geom()`

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20,            
                 color="black",      
                 fill="light blue")  
```

## Modifying a geometric object by changing *aes()*

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           fill = GENDER)) +
  geom_histogram(bins=20, 
                 color="grey30")
```

## Geometric Objects: geom-density()

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_density()           
```

The code chunk below plots two kernel density lines by using *colour* or *fill* arguments of *aes()*

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS, 
           colour = GENDER)) +
  geom_density()
```

## **Geometric Objects: geom_boxplot**

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS,       
           x= GENDER)) +    
  geom_boxplot()            
```

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot(notch=TRUE)
```

## Geometric Objects: geom_violin

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_violin()
```

## **Geometric Objects: geom_point()**

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point()            
```

## ***Geom*** **Objects can be combined**

The code chunk below plots the data points on the boxplots by using both `geom_boxplot()` and `geom_point()`.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot() +                    
  geom_point(position="jitter", 
             size = 0.5)        
```

## **Essential Grammatical Elements in ggplot2: stat**

The [Statistics functions](https://ggplot2.tidyverse.org/reference/index.html#stats) statistically transform data, usually as some form of summary. For example:

-   frequency of values of a variable (bar graph)

    -   a mean

    -   a confidence limit

-   There are two ways to use these functions:

    -   add a `stat_()` function and override the default geom, or

    -   add a `geom_()` function and override the default stat.

## **Working with `stat()`**

The boxplots below are incomplete because the positions of the means were not shown.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot()
```

## **Working with stat - the *stat_summary()* method**

The code chunk below adds mean values by using [`stat_summary()`](https://ggplot2.tidyverse.org/reference/stat_summary.html) function and overriding the default geom.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",       
               fun = "mean",         
               colour ="red",        
               size=4)               
```

## **Working with stat - the `geom()` method**

The code chunk below adding mean values by using `geom_()` function and overriding the default stat.

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  geom_point(stat="summary",        
             fun="mean",           
             colour="red",          
             size=4)          
```

## **Adding a best fit curve on a scatterplot?**

In the code chunk below, [`geom_smooth()`](https://ggplot2.tidyverse.org/reference/geom_smooth.html) is used to plot a best fit curve on the scatterplot.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(size=0.5)
```

The default smoothing method can be overridden as shown below.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5)
```

## **Essential Grammatical Elements in ggplot2: Facets**

Facetting generates small multiples (sometimes also called trellis plot), each displaying a different subset of the data. They are an alternative to aesthetics for displaying additional discrete variables. ggplot2 supports two types of factes, namely: [`facet_grid()`](https://ggplot2.tidyverse.org/reference/facet_grid.html) and [`facet_wrap`](https://ggplot2.tidyverse.org/reference/facet_wrap.html).

### **Working with `facet_wrap()`**

[`facet_wrap`](https://ggplot2.tidyverse.org/reference/facet_wrap.html) wraps a 1d sequence of panels into 2d. This is generally a better use of screen space than facet_grid because most displays are roughly rectangular.

The code chunk below plots a trellis plot using `facet-wrap()`.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)
```

### **`facet_grid()` function**

[`facet_grid()`](https://ggplot2.tidyverse.org/reference/facet_grid.html) forms a matrix of panels defined by row and column facetting variables. It is most useful when you have two discrete variables, and all combinations of the variables exist in the data.

The code chunk below plots a trellis plot using `facet_grid()`.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_grid(~ CLASS)
```

## **Essential Grammatical Elements in ggplot2: Coordinates**

The *Coordinates* functions map the position of objects onto the plane of the plot. There are a number of different possible coordinate systems to use, they are:

```         
-   [`coord_cartesian()`](https://ggplot2.tidyverse.org/reference/coord_cartesian.html): the default cartesian coordinate systems, where you specify x and y values (e.g. allows you to zoom in or out).
-   [`coord_flip()`](https://ggplot2.tidyverse.org/reference/coord_flip.html): a cartesian system with the x and y flipped.
-   [`coord_fixed()`](https://ggplot2.tidyverse.org/reference/coord_fixed.html): a cartesian system with a "fixed" aspect ratio (e.g. 1.78 for a "widescreen" plot).
-   [`coord_quickmap()`](https://ggplot2.tidyverse.org/reference/coord_map.html): a coordinate system that approximates a good aspect ratio for maps.
```

### **Working with Coordinate**

By the default, the bar chart of ggplot2 is in vertical form.

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```

The code chunk below flips the horizontal bar chart into vertical bar chart by using `coord_flip()`.

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip()
```

### **Changing the y- and x-axis range**

The scatterplot on the right is slightly misleading because the y-aixs and x-axis range are not equal.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5)
```

The code chunk below fixed both the y-axis and x-axis range from 0-100.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
```

## **Essential Grammatical Elements in ggplot2: themes**

Themes control elements of the graph not related to the data. For example:

-   background colour

-   size of fonts

-   gridlines

-   colour of labels

Built-in themes include: - `theme_gray()` (default) - `theme_bw()` - `theme_classic()`

A list of theme can be found at this [link](https://ggplot2.tidyverse.org/reference/ggtheme.html). Each theme element can be conceived of as either a line (e.g. x-axis), a rectangle (e.g. graph background), or text (e.g. axis title).

### **Working with theme**

The code chunk below plot a horizontal bar chart using `theme_gray()`.

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()
```

A horizontal bar chart plotted using `theme_classic()`.

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()
```

A horizontal bar chart plotted using `theme_minimal()`.

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()
```

## **Reference**

-   Hadley Wickham (2023) [**ggplot2: Elegant Graphics for Data Analysis**](https://ggplot2-book.org/). Online 3rd edition.

-   Winston Chang (2013) [**R Graphics Cookbook 2nd edition**](https://r-graphics.org/). Online version.

-   Healy, Kieran (2019) [**Data Visualization: A practical introduction**](https://socviz.co/). Online version

-   [Learning ggplot2 on Paper – Components](https://henrywang.nl/learning-ggplot2-on-paper-components/)

-   [Learning ggplot2 on Paper – Layer](https://henrywang.nl/learning-ggplot2-on-paper-layer/)

-   [Learning ggplot2 on Paper – Scale](https://henrywang.nl/tag/learning-ggplot2-on-paper/)
